\MakeUppercase{Layered $H$-Partitions of $k$-Planar Graphs} \thanks{This work was partly funded by NSERC and MRI.}

### 2019-05-07 k-planar.tex
This is working cleanly

  \item \label{ancestor-boundary} For each non-root node $x\in V(T)$, there exists a partition of $V(F_x)$ into $g\le 3$ sets $P_1,\ldots,P_g$ such that, for each $i\in\{1,\ldots,g\}$, $F[P_i]$ is a path that is contained in $Y_a$ for some $K$-ancestor $a$ of $x$ \emph{or} $F[P_i]$ is a path in $F_{r'}$.

@@

  \item Something else?
\end{compactenum}

Properties~(PR1)--(PR3) are defining properties of a PS-tree, and they imply the remaining properties.

@@

We claim that the curve $C$ begins in $F_x\subseteq F^x_a\subseteq F_a$, remains in the interior of $F_a$, and eventually reaches the boundary $F^{\bar{x}}_a\cap F^x_a$.  The fact that $C$ remains in $F_a$ is due to the fact that, if it crossed the boundary of $F_a$, then one of $vw$ or $wz$ would intersect $Y_{a'}$ for some ancestor $a'$ of $a$.  If $vw$ intersects $Y_{a'}$, then $v\not\in S_x$.  If $wz$ intersects $Y_{a'}$ then $w,z\not\in S_a$. The fact that $C$ eventually reaches follows from the fact that $w\in Y_a$ or $wz$ contains a point of $Y_a$.

Therefore, to upper-bound $|B_x|$, it suffices to upper-bound the number of faces $F^{\bar{x}}_a$ of $A$ that can be reached with curves that begin in $F_x$, remain in the interior of $F_a$, and terminate in $F^{\bar{x}}_a$ without intersecting more than $2k$ edges in $A$.  To avoid triple subscripts we say that a curve visits $a_i$ if it visits $F^{\bar{x}}_{a_i}$.  We can further restrict ourselves to studying curves that cross the fewest number of edges of $A$ since if a curve exists that crosses at most $k$ edges then the curve that crosses the fewest number of possible edges also crosses at most $k$ edges.  We will call such a curve a \emph{lightest} curve from $x$ to $a$.

Again, we let $a_0,\ldots,a_r$ denote the vertices of the path, in $K$ from the root $a_0$ to $x=a_r$.  We claim that, there exists a lightest curve from $a_r$ to $a_0$ that is monotonic in the sense that the curve visits $a_{i_0},\ldots,a_{i_t}$ in this order, where $i_0=r > i_1>\cdots>i_t=0$.
This follows from (PR\ref{ancestor-boundaries}):  For any node $a_i$, $i>0$, there are at most three faces of $A$ that share a part of their boundary with $F_{a_i}$.  Each of these faces also necessarily shares a part of its boundary with $F^{\bar{x}}_{a_i}$.  Therefore it is never advantageous for a path to go from $a_i$ to $a_j$ for some $j>i$ since the path will eventually have to return to one of these three faces.

At this point we are essentially done.  Any monotonic curve that visits $a_i$ has at most 3 choices for the next face $a_j$ $j<i$ to visit.  Therefore, the number of faces of $A$ that can be reached from $F_x$ by admissible curves that cross $2k$ edges of $A$ is at most $3^{2k}$. Therefore,
\[
   |B_x| \le \sum_{j=0}^2k 9^j = (9^{2k+1}-1)/8 \enspace .
\]

This completes the proof

\begin{thm}
    Every $k$-planar graph $G$ has a $(k9^{2k+1}/8,3k)$-partition.
\end{thm}

>>>>

  \item \label{ancestor-boundary} For each non-root node $x\in V(T)$, there exists a partition of $V(F_x)$ into $g\le 3$ sets $P_1,\ldots,P_g$ such that, for each $i\in\{1,\ldots,g\}$, $F[P_i]$ is a path that is contained in $Y_a$ for some $K$-ancestor $a$ of $x$ \emph{or} $F[P_i]$ is a path or cycle in $F_{r'}$.

@@

  % \item Something else?
\end{compactenum}

Properties~(PR1)--(PR3) are defining properties of a PS-tree. Property~(PR4)

@@

We claim that the curve $C$ begins in $F_x\subseteq F^x_a\subseteq F_a$, remains in the interior of $F_a$, and eventually reaches the boundary $F^{\bar{x}}_a\cap F^x_a$.  The fact that $C$ remains in $F_a$ is due to the fact that, if it crossed the boundary of $F_a$, then one of $vw$ or $wz$ would intersect $Y_{a'}$ for some ancestor $a'$ of $a$.  If $vw$ intersects $Y_{a'}$, then $v\not\in S_x$.  If $wz$ intersects $Y_{a'}$ then $w,z\not\in S_a$. The fact that $C$ eventually reaches follows from the fact that $w\in Y_a$ or $wz$ contains a point of $Y_a$.  We call such a curve an \emph{admissible} curve.

Therefore, to upper-bound $|B_x|$, it suffices to upper-bound the number of faces $F^{\bar{x}}_a$ of $A$ that can be reached with curves that begin in $F_x$, remain in the interior of $F_a$, and terminate in $F^{\bar{x}}_a$ without intersecting more than $2k$ edges in $A$.  To avoid triple subscripts we say that a curve visits $a_i$ if it visits $F^{\bar{x}}_{a_i}$.  We can further restrict ourselves to studying curves that cross the fewest number of edges of $A$ since if a curve exists that crosses at most $k$ edges then the curve that crosses the fewest number of possible edges also crosses at most $k$ edges.  We will call such a curve a \emph{lightest} admissible curve from $x$ to $a$.

Again, we let $a_0,\ldots,a_r$ denote the vertices of the path, in $K$ from the root $a_0$ to $x=a_r$.  We claim that, there exists a lightest admissible curve from $a_r$ to $a_0$ that is monotonic in the sense that the curve visits $a_{i_0},\ldots,a_{i_t}$ in this order, where $i_0=r > i_1>\cdots>i_t=0$.
This follows from (PR\ref{ancestor-boundaries}):  For any node $a_i$, $i>0$, there are at most three faces of $A$ that share a part of their boundary with $F_{a_i}$.  Each of these faces also necessarily shares a part of its boundary with $F^{\bar{x}}_{a_i}$.  Therefore no lightest admissible curve ever $a_i$ before $a_j$ for some $j>i$ since the curve will eventually have to return to one of these three faces.

At this point we are essentially done.  Any lightest admissible curve that visits $a_i$ has at most 3 choices for the next face $a_j$ $j<i$ to visit.  Therefore, the number of faces of $A$ that can be reached from $F_x$ by admissible curves that intersect at most $\ell$ edges of $A$ is at most $3^{\ell}$. Therefore,
\[
   |B_x| \le \sum_{\ell=0}^{2k+1} 3^\ell = (3^{2k+2}-1)/2 
       = (9^{k+1}-1)/2 \enspace .
\]

This is all we need to complete the proof of:

\begin{thm}
    Every $k$-planar graph $G$ has a $((9^{k+1}-1)/2-1,3k^2)$-partition.
\end{thm}

\begin{proof}
  The $H$-partition $(B_x:x\in V(H))$ of $G$ described above has treewidth $(9^{k+1}-1)/2-1$.  For each $x\in V(H)$ and each integer $i$, $|B_x\cap V_i|\le 3k$. \Eqref{distance-preserving} implies setting $V_i' = V_{ki}\cup\cdots\cup V_{k(i+1)-1}$ yields a layering $(V_0',V_1',\ldots)$ of  $G$ and $|B_x\cap V_i'|\le 3k^2$ for all integers $i\ge 0$ and all $x\in V(H)$.
\end{proof}
